---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loads all the 

```{r globalresources}
# loads all libraries used in the research
library(knitr)
library(lattice)
# sets the option echo = TRUE for all the code chunks
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

```{r loadpreprocess}
## 1. Load the data (i.e. read.csv())

# temporary unzip activity.zip archive
temp <- unzip("activity.zip")
# reads the dataset with read.csv function
data <- read.csv(temp)
# removes the temporary .csv file
unlink(temp)


## 2. Process/transform the data (if necessary) into a format suitable for your
## analysis

# transforms date values from Factor to Date
data$date <- as.Date(as.character(data$date)) 


## (3.) Sample

# here's a preview of the data dataframe
head(data)
```

## What is mean total number of steps taken per day?

```{r meantotalnum}
## 1. Calculate the total number of steps taken per day

# calculates the total number of steps per day
tot_steps_per_day <- aggregate(data$step, by=list(data$date), FUN=sum)
# changes the Group.1 variable name to step
colnames(tot_steps_per_day) <- c("date", "steps")


## 2. If you do not understand the difference between a histogram and a barplot,
## research the difference between them. Make a histogram of the total number 
## of steps taken each day
hist(tot_steps_per_day$steps, main = "Total Number of Steps Taken Each Day", 
     xlab = "Steps", col = "red")


## 3. Calculate and report the mean and median of the total number of steps
## taken per day
mean <- format(round(mean(tot_steps_per_day$steps, na.rm = T), 2), nsmall = 2)
median <- median(tot_steps_per_day$steps, na.rm = T)
```

### Total number of steps taken per day:

Mean: `r mean`

Median: `r median`

## What is the average daily activity pattern?

```{r avgdailyactivitypattern}
## 1. Make a time series plot (i.e. type = "l") of the 5-minute interval
## (x-axis) and the average number of steps taken, averaged across all days 
## (y-axis)

# aggregates steps by interval, using the mean function, and creates the plot data
plot_data <- aggregate(data$step, by=list(data$interval), FUN="mean", na.rm = T)
# renames the two columns, so they can be more readable
colnames(plot_data) <- c("interval", "avg_steps")
# creates the required time series plot
with(plot_data, plot(interval, avg_steps, type = "l", 
                     main = "Average Number of Steps Taken", 
                     xlab = "5-minute Interval", 
                     ylab = "Average Number of Steps"))


## 2. Which 5-minute interval, on average across all the days in the dataset, 
## contains the maximum number of steps?

# finds out the row that has the max average steps
interval_row <- which.max(plot_data$avg_steps)
# gets the interval value for that observation
interval_max_steps <- plot_data[interval_row, ]$interval
```

Time interval that contains the max number of steps: `r interval_max_steps`


## Imputing missing values

```{r missinvalues}
## 1. Calculate and report the total number of missing values in the dataset 
## (i.e. the total number of rows with NAs)
sum(!complete.cases(data))


## 2. Devise a strategy for filling in all of the missing values in the dataset.
## The strategy does not need to be sophisticated. For example, you could use 
## the mean/median for that day, or the mean for that 5-minute interval, etc.

# first I need to know which variables are affected by missing values
na_cols <- cbind(sum(is.na(data$steps)), sum(is.na(data$date)), 
                 sum(is.na(data$interval)))
colnames(na_cols) <- c("steps", "date", "interval")
na_cols
```

Only steps variable has NA values. I assume that I can use the mean for that 5-minute interval, calculated using the rest of observations

```{r missfill}
## 3. Create a new dataset that is equal to the original dataset but with the 
## missing data filled in.

# first converts the two date variable into character to avoid row binding errors
data$date <- as.character(data$date)
# them copies the original data
data_miss <- data[!complete.cases(data), ]
# after that calculates the mean for each interval
interval_mean <- aggregate(data$step, by=list(data$interval), FUN="mean", 
                           na.rm=TRUE)
# changes the interval_mean variable names
colnames(interval_mean) <- c("interval", "steps")

# then replaces the NA values with the corresponding interval mean
data_miss <- merge(data_miss, interval_mean, by="interval")
data_miss <- data_miss[order(data_miss$date, data_miss$interval),]
head(data_miss)

# gets the number of steps from the interval_mean dataframe, then drops all 
# variables except for steps, date and interval
data_miss$steps <- data_miss$steps.y
data_miss <- data_miss[, c(5, 3, 1)]

# append the two dataframes and converts date variable into Date
data_fill <- rbind(data_miss, data[complete.cases(data), ])
data$date <- as.Date(data_fill$date)
data_fill$date <- as.Date(data_fill$date)

# finally shows a preview of the filled dataframe
head(data_fill)


## 4.Make a histogram of the total number of steps taken each day and Calculate 
## and report the mean and median total number of steps taken per day. Do these 
## values differ from the estimates from the first part of the assignment? What 
## is the impact of imputing missing data on the estimates of the total daily 
## number of steps?

# calculates the total number of steps per day
tot_steps_per_day_fill <- aggregate(data_fill$step, by=list(data_fill$date), 
                                    FUN=sum)
# changes the Group.1 variable name to step
colnames(tot_steps_per_day_fill) <- c("date", "steps")
# creates the required histogram
hist(tot_steps_per_day_fill$steps, main = "Total Number of Steps Taken Each Day", 
     xlab = "Steps", col = "red")
# calculates the mean and median total number of steps taken per day
mean_fill <- format(round(mean(tot_steps_per_day_fill$steps), 2), nsmall = 2)
median_fill <- format(round(median(tot_steps_per_day_fill$steps), 2), 
                      nsmall = 0)
```

### Total number of steps taken per day:

Mean: `r mean_fill`

Median: `r median_fill`

After the replacement of NA values with the mean value grouped by interval, the mean remained the same. That's because of the tecnique used to fill the missing values (the mean value indeed). 
The median otherwise raised up of a unit (10766.19 instead of 10765), and that's because we introduced more valid observations for the variable step (exactly 2304).

## Are there differences in activity patterns between weekdays and weekends?

```{r diffweekdaysweekends}
## 1. Create a new factor variable in the dataset with two levels - "weekday" 
## and "weekend" indicating whether a given date is a weekday or weekend day.

# gets the days of the weekend with the correct position
weekend <- weekdays(seq(as.Date("2012/10/6"), as.Date("2012/10/7"), "days"))
# gives to the weekday_ind the value "weekend" by default
data_fill$weekday_ind <- "weekday"
# gives to weekend days the appropriate weekday_ind value
data_fill[weekdays(data_fill$date) %in% weekend, ]$weekday_ind <- "weekend"
# transforms the former variable to factor
data_fill$weekday_ind <- as.factor(data_fill$weekday_ind)


## 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 
## 5-minute interval (x-axis) and the average number of steps taken, averaged 
## across all weekday days or weekend days (y-axis). See the README file in the 
## GitHub repository to see an example of what this plot should look like using 
## simulated data.

# creates the plot dataframe, aggregating steps
plot_data_fill <- with(data_fill, 
                       aggregate(steps, by=list(weekday_ind, interval), 
                                 FUN="mean"))
# replaces the data frame names with the correct ones
colnames(plot_data_fill) <- c("weekday_ind", "interval", "steps")

# creates the required chart
xyplot(steps ~ interval | weekday_ind, layout = c(1, 2), 
       data=plot_data_fill, type='l')
```